|

### 2017-10-06 manuscript.Rmd
Finish Daniel's edits (#2) and adds Patrick's suggested changes.
@eddelbuettel: I think this is ready to go on my end, just give me a thumbs up if it's okay to submit!

## Introduction

@@

Docker is a popular open source tool to create, distribute, deploy,
and run software applications using *containers*.  Containers provide a
virtual environment (see \citet{Clark2014} for an overview of common virtual environments) requiring all operating-system components an application needs to run:
code, runtime, system tools, system runtime.  Docker containers are
lightweight as they share the operating system kernel, starting instantly
using a layered filesystem which minimizes disk footprint and download time,
are built on open standards that run on all major platforms (Linux, Mac,
Windows), and provide an added layer of security by running an application
in an isolated environment \citep{what-docker}.  Familiarity with a few key
terms is helpful in understanding this paper. The term "container" refers to 
an isolated software environment on a computer. R users can think of running
a container as analgous to loading an R package; it makes a set of software
functions available.  A Docker "image" is a binary archive of that software,
analgous to an R binary package -- a given version need be downloaded only
once and then can be "run" to create a container whenever it is needed.  
A "Dockerfile" is a recipe, the source-code, to create a Docker image.  Development
and contributions to the Rocker project focuses on the construction, organization
and maintenance of these Dockerfiles.

## Design principles & use cases

@@

      While most virtual computing environments (AMI images, scientific cloud
      platforms like cyverse, etc) provide some subset of these, they are most
      effective in concert.

- **Portable**: from laptop to cloud
- **Interfaces**: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
- **Sandboxed**: experiment without touching local system
- **Transparent**: Dockerfiles & automated builds provide a transparent and verifiable record of precisely what is installed.
- **Optimized**: community-tweaked configurations
- **Extensible**: interactively and persistent community extensions
- **Versioned**: reproducible environments

@@

For instance, the following code installs Docker on almost any Linux-based server and
launches a Rocker container providing the RStudio-server environment
over a web interface.

    wget -qO- https://get.docker.com/ | sh 
    sudo docker run -d -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio

Many academic and commercial cloud providers make it
possible to execute such code snippets when a container is launched, without
ever needing to `ssh` into the machine. The user may log into the server merely
by pasting its IP address or DNS name (followed by the chosen port, RStudio's default `:8787` in the above example)
into a browser and entering the
appropriate password.  This provides the user with a familiar,
interactive environment running on a remote machine while requiring
a minimum of expertise.

@@

Users can easily determine the software stack installed on any Rocker image
by examining the associated Dockerfile recipe, which provides a concise, human-readable
record of the installation.
All Rocker images use automated builds through Docker Hub, which also acts as the central,
default repository distributing the images.  Using automated builds rather than uploading
pre-built image binaries to the Docker Hub avoids the potential for the build
not to match the recipe.  The corresponding Dockerfile is visible both on the Docker Hub
and in the linked GitHub repository of the Rocker project (https://github.com/rocker-org), 
which provides a transparent versioned history of all changes made to these recipes,
as well as documentation, a community wiki, and issue trackers for discussing proposed
changes, bugs, improvements to the Dockerfiles and troubleshoot any issues users may encounter.

@@

The versioned stack, (`r-ver`, `rstudio`, `tidyverse`, `verse`,
and `geospatial`), provides images which are intended to build an identical
software stack every time, regardless of the release of new libraries and packages.  This 
can be important for users who need computational reproducibility more than having the latest
release of any piece of software, since subsequent releases can alter the behavior of code,
introducing errors or altering previous results.  Users should specify an R version tag in
the Docker image name to request a version stable image, e.g. `rocker/verse:3.4.0`.  If no tag
is explicitly requested, Docker will provide the image with the tag `:latest`, which will
always have the latest available versions of the software (built nightly).

@@

## Rocker organization and workflow

@@

The Rocker project aims to provide a small core of Docker images that
serve as convenient 'base' images on which other users can build custom
R environments by writing their own Dockerfiles, while also providing a
'batteries included' approach of images that can be used out of the box.
The challenges of balancing diverse needs driven by very different use
cases (discussed in more detail below) against the overarching goals
of creating images that are still sufficiently light-weight, easy to use
and easy to maintain is a difficult art.  The implementation in both
individual Rocker images and image stacks can never perfect that balance
for everyone, but today reflects the considerable community input and
testing over past few years.

@@

## The `debian:testing`-based images

@@

The `debian:testing`-based stack currently includes seven images
actively maintained by the Rocker development team (Table 1).
`r-base` builds on `debian:testing`, and the other six in the stack
each build directly from `r-base`.  The `r-base` image is unique in
that it is designated as the official image for the R language by the
Docker organization itself.  This official image is reviewed and then
built by employees of Docker Inc based on a Dockerfile maintained by
the Rocker team.  Consequently, users should refer to this image in
Docker commands without an organization namespace, e.g. `docker run -t
r-base` to access the official image.  All other images in the Rocker
project are not individually reviewed and built by Docker Inc and must
be referenced using the `rocker` namespace,
e.g. `docker run -t rocker/r-devel`.

@@

As these images focus on developers and/or as base images for custom uses,
this stack does not include many specific R packages. Additional dependencies
and packages can easily be installed from `apt`.  R packages not available
in the `apt` repositories can be installed directly from CRAN using the
`littler` scripts, see *Extending Rocker images* section below.

@@

image            | description                               | size   | downloads 
---------------- | ----------------------------------------- | ------ | ------- 
[r-base](https://hub.docker.com/r/_/r-base)        | official image with current version of R  | 254 MB | 510,000   
[r-devel](https://hub.docker.com/r/rocker/r-devel) |  R-devel added side-by-side r-base (using alias `RD`)    | 1 GB | 4,000
[drd](https://hub.docker.com/r/rocker/drd) |  lightweight r-devel, built almost daily  | 571 MB | 4,000
[r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanitizers | 1.1 GB  | 1,000  
[r-devel-ubsan-clang](https://hub.docker.com/r/rocker/r-devel-ubsan-clang)            |  Sanitizers, clang c compiler (instead of gcc) | 1.1 GB | 525
[rstudio:testing](https://hub.docker.com/r/rocker/r-devel-san)        |  rstudio on debian:testing | 1.1 GB | 1,000 
[shiny](https://hub.docker.com/r/rocker/shiny)     |  shiny-server on r-base               | 409 MB | 63,000

@@

This stack emphasizes stability and reproducibility of the Docker build.
This stack was introduced much more recently (November 2016) 
in response to considerable user input and requests. The key
feature of this stack is the ability to run older versions of R along
with the then-contemporaneous versions of R packages.  A user specifies the
version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
refer to an image with R version 3.3.1 installed.  Omitting the tag
is equivalent to using the tag `latest`, which, as the name implies,
will always point to an image using the current R release.  Thus, users
wanting to create downstream Dockerfiles which are based on the current
release at time (but will continue to reconstruct the same environment in
the future after newer R versions are released) should explicitly include
the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
of writing, and not the `latest` tag.  Users can also run the current
development version of R using the tag `devel`, which is built nightly
from R-devel sources from `subversion`.

@@

**Version tags**:  The version tags are propagated throughout this
stack: _e.g._, `rocker/tidyverse:devel` will provide the currently
released versions of the R packages in the `tidyverse` \citep{tidyverse}
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
starting 2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accommodate changing dependencies.  Dockerfiles for
past versions (e.g. < `3.4.0` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accommodate new features or dependencies.

**Installation**: In this stack, the desired version of R is always built
directly from source rather than the `apt` repositories.  Compilers and
dependencies are still installed from the stable `apt` repositories,
and thus lag behind the more recent versions found in the `testing`
stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
code-named `jessie`, while `3.4.0`, `devel`, and `latest` are based on
Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much
newer versions of common system dependencies and compilers.  Dependencies
needed to compile R that are not required at runtime are removed once R is
installed, keeping the base images light-weight for faster download times.
While most system dependencies required by common R packages can still
be installed from the `apt` repositories, occasionally a more recent
version must be compiled from source (e.g. the Gibbs Sampling library,
JAGS, and the geospatial toolkit GDAL, must both be compiled from source
on `debian:jessie` images).  In this stack, users should avoid installing R
packages using `apt` without careful consideration, as this will install a second, 
probably different version
of R from the Debian repositories and a dated version of the R package
(since any `r-cran-pkgname` package in the Debian repositories will
depend on `r-base` in `apt` as well).

@@

tag     | apt repos | MRAN date    | Build frequency  | images with tag 
--------|-----------|--------------|------------------|-------------------------------           
devel   | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
latest  | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.4.1   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.4.0   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.3.3   | `jessie`  | 2017-04-21   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial` 
3.3.2   | `jessie`  | 2017-03-06   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.3.1   | `jessie`  | 2016-10-31   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.3.0   | `jessie`  | 2016-06-21   | monthly          | `r-ver`
3.2.0   | `jessie`  | 2015-06-18   | monthly          | `r-ver`
3.1.0   | `jessie`  | 2014-09-17   | monthly          | `r-ver`

@@

## Maintenance and Updates

@@

## License ##

@@

<!-- 
Use cases: instruction, development, 
-->

## Conclusions

Over the past several years, Docker has seen immense adoption across industry and academia. 
The Open Container initiative \citep{oci} now provides an open standard that has further
extended this container approach to research environments through projects such as 
Singularity \citep{singularity}, used on clusters without root access, and Shifter \citep{shifter}
for use on leadership high performance computing facilities. Containerization promises
to solve numerous challenges such as portability and replicability in research computing,
which often relies on complex and heterogeneous software stacks \citep{Boettiger2015}. Yet
implementing such environments in containers is not a trivial task, and not all implementations
provide the same usability, portability or reproducibility.  Here we have detailed the 
approach taken by the Rocker project in creating and maintaining these environments through
an open and community-driven process.  This structure of the Rocker project has evolved
over three years of operation while drawing in an ever widening base of academic researchers, 
university instructors and industry users.  We believe this overview will be instructive
not only to users and developers interested in the Rocker project, but as a model for
similar efforts around other environments or domains.

>>>>

# Introduction

@@

Docker is a popular open source tool to create, distribute, deploy,
and run software applications using *containers*.  Containers provide a
virtual environment (see \citet{Clark2014} for an overview of common virtual environments) requiring all operating-system components an application needs to run:
code, runtime, system tools, system runtime.  Docker containers are
lightweight as they share the operating system kernel, starting instantly
using a layered filesystem which minimizes disk footprint and download time,
are built on open standards that run on all major platforms (Linux, Mac,
Windows), and provide an added layer of security by running an application
in an isolated environment \citep{what-docker}.  Familiarity with a few key
terms is helpful in understanding this paper. The term "container" refers to 
an isolated software environment on a computer. R users can think of running
a container as analogous to loading an R package; it makes a set of software
functions available.  A Docker "image" is a binary archive of that software,
analogous to an R binary package -- a given version need be downloaded only
once and then can be "run" to create a container whenever it is needed.
A "Dockerfile" is a recipe, the source-code, to create a Docker image.  Development
and contributions to the Rocker project focuses on the construction, organization
and maintenance of these Dockerfiles.

# Design principles & use cases

@@

- **Portable**: from laptop to cloud
- **Interfaces**: Console-based (R, bash, etc) or RStudio
- **Sandboxed**: experiment without touching local system
- **Transparent**: Dockerfiles & automated builds provide a transparent and verifiable record of precisely what is installed.
- **Optimized**: community-tweaked configurations
- **Extensible**: interactively and persistent community extensions
- **Versioned**: reproducible environments

@@

For instance, the following bash commands install Docker on almost any 
Linux-based server and then launch a Rocker container providing the 
RStudio-server environment over a web interface.

    wget -qO- https://get.docker.com/ | sh 
    sudo docker run -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio

The `docker run` option `-p` sets the port on which RStudio will appear, 8787 is
the RStudio default. Many academic and commercial cloud providers make it
possible to execute such code snippets when a container is launched, without
ever needing to `ssh` into the machine. The user may log into the server merely
by pasting its IP address or DNS name (followed by the chosen port, e.g. `:8787`)
into a browser and entering the
appropriate password.  This provides the user with a familiar,
interactive environment running on a remote machine while requiring
a minimum of expertise.

@@

Users can easily determine the software stack installed on any Rocker image
by examining the associated Dockerfile recipe, which provides a concise, human-readable
record of the installation.
All Rocker images use automated builds through Docker Hub, which also acts as the central,
default repository distributing the images.  Using automated builds rather than uploading
pre-built image binaries to the Docker Hub avoids the potential for the build
not to match the recipe.  The corresponding Dockerfile is visible both on the Docker Hub
and in the linked GitHub repository of the Rocker project (https://github.com/rocker-org), 
which provides a transparent versioned history of all changes made to these recipes,
as well as documentation, a community wiki, and issue trackers for discussing proposed
changes, bugs, improvements to the Dockerfiles and troubleshoot any issues users may encounter.
Having these public source files built automatically by a trusted provider (Docker Hub),
rather than built locally and uploaded only in binary form provides is also useful from a
security perspective in avoiding malware.

@@

Access to specific versions of software  
can be important for users who need computational reproducibility more than having the latest
release of any piece of software, since subsequent releases can alter the behavior of code,
introduce errors or otherwise alter previous results.  
The versioned stack, (`r-ver`, `rstudio`, `tidyverse`, `verse`,
and `geospatial`), provides images which are intended to build an identical
software stack every time, regardless of the release of new libraries and packages.  
Users should specify an R version tag in
the Docker image name to request a version stable image, e.g. `rocker/verse:3.4.0`.  If no tag
is explicitly requested, Docker will provide the image with the tag `:latest`, which will
always have the latest available versions of the software (built nightly).

@@

# Rocker organization and workflow

@@

The Rocker project aims to provide a small core of Docker images that
serve as convenient 'base' images on which other users can build custom
R environments by writing their own Dockerfiles, while also providing a
'batteries included' approach of images that can be used out of the box.
The challenges of balancing diverse needs driven by very different use
cases against the overarching goals
of creating images that are still sufficiently light-weight, easy to use
and easy to maintain is a difficult art.  The implementation in both
individual Rocker images and image stacks can never perfect that balance
for everyone, but today reflects the considerable community input and
testing over past few years.

@@

### The `debian:testing`-based images

@@

The `debian:testing`-based stack currently includes seven images
actively maintained by the Rocker development team (Table 1).
`r-base` builds on `debian:testing`, and the other six in the stack
each build directly from `r-base`.  The `r-base` image is unique in
that it is designated as the official image for the R language by the
Docker organization itself.  This official image is reviewed and then
built by employees of Docker Inc based on a Dockerfile maintained by
the Rocker team.  Consequently, users should refer to this image in
Docker commands without an organization namespace, e.g. `docker run -t
r-base` to access the official image.  All other images in the Rocker
project are not individually reviewed and built by Docker Inc and must
be referenced using the `rocker` namespace,
e.g. `docker run -ti rocker/r-devel`.

@@

As these images focus on developers and/or as base images for custom uses,
this stack does not include many specific R packages. Additional dependencies
and packages can easily be installed from `apt`.  R packages not available
in the `apt` repositories can be installed directly from CRAN using the
`littler` scripts, as described in <https://rocker-project.org/use>.

@@

image            | description                               | size   | downloads 
---------------- | ----------------------------------------- | ------ | ------- 
[r-base](https://hub.docker.com/r/_/r-base)        | official image with current version of R  | 254 MB | 632,000   
[r-devel](https://hub.docker.com/r/rocker/r-devel) |  R-devel added side-by-side r-base (using alias `RD`)    | 1 GB | 4,000
[drd](https://hub.docker.com/r/rocker/drd) |  lightweight r-devel, built weekly  | 571 MB | 4,000
[r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanitizers | 1.1 GB  | 1,000  
[r-devel-ubsan-clang](https://hub.docker.com/r/rocker/r-devel-ubsan-clang)            |  Sanitizers, clang c compiler (instead of gcc) | 1.1 GB | 525
[rstudio:testing](https://hub.docker.com/r/rocker/r-devel-san)        |  rstudio on debian:testing | 1.1 GB | 1,000 
[shiny](https://hub.docker.com/r/rocker/shiny)     |  shiny-server on r-base               | 409 MB | 123,000

@@

This stack emphasizes stability and reproducibility of the Docker build.
This stack was introduced much more recently (November 2016) 
in response to considerable user input and requests. The key
feature of this stack is the ability to run older versions of R along
with the then-contemporaneous versions of R packages.  A user specifies the
version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
refer to an image with R version 3.3.1 installed.  Omitting the tag
is equivalent to using the tag `latest`, which, as the name implies,
will always point to an image using the current R release.  Thus, users
wanting to create downstream Dockerfiles which are based on the current
release at time (but will continue to reconstruct the same environment in
the future after newer R versions are released) should explicitly include
the corresponding version tag, e.g. `rocker/r-ver:3.4.2` at the time
of writing, and not the `latest` tag.  Users can also run the current
development version of R using the tag `devel`, which is built nightly
from R-devel sources from `subversion`.

@@

**Version tags**:  The version tags are propagated throughout this
stack: _e.g._, `rocker/tidyverse:devel` will provide the currently
released versions of the R packages in the `tidyverse` \citep{tidyverse}
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
starting 2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accommodate changing dependencies.  Dockerfiles for
past versions (e.g. < `3.4.0` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accommodate new features or dependencies.
Version tags also obey semantics, so that omitting the second or third
position of the tag is identical to asking for the most recent version:
i.e. `rocker/verse:3.3` is the same as `rocker/verse:3.3.3`, and 
`rocker/verse:3` is (at the time of writing), `rocker/verse:3.4.2`.
This is accomplished using post-build hooks in Docker Hub, see examples at
<https://github.com/rocker-org/rocker-versioned/> for details.

**Installation**: In this stack, the desired version of R is always built
directly from source rather than the `apt` repositories.  Compilers and
dependencies are still installed from the stable `apt` repositories,
and thus lag behind the more recent versions found in the `testing`
stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
code-named `jessie`, while `3.4.0` - `3.4.2`, `devel`, and `latest` are based on
Debian 9.0, `stretch`, (released 2017-06-17, while R was at `3.4.0`), 
and thus have access to much
newer versions of common system dependencies and compilers.  Dependencies
needed to compile R that are not required at runtime are removed once R is
installed, keeping the base images light-weight for faster download times.
While most system dependencies required by common R packages can still
be installed from the `apt` repositories, occasionally a more recent
version must be compiled from source (e.g. the Gibbs Sampling library,
JAGS, and the geospatial toolkit GDAL, must both be compiled from source
on `debian:jessie` images).  In this stack, users should avoid installing R
packages using `apt` without careful consideration, as this will install a second, 
probably different version
of R from the Debian repositories and a dated version of the R package
(since any `r-cran-pkgname` package in the Debian repositories will
depend on `r-base` in `apt` as well).

@@

tag     | apt repos | MRAN date    | Build frequency  | images with tag 
--------|-----------|--------------|------------------|-------------------------------           
devel   | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
latest  | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.4.2   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.4.1   | `stretch` | 2017-09-28 | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.4.0   | `stretch` | 2017-06-30 | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.3.3   | `jessie`  | 2017-04-21   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial` 
3.3.2   | `jessie`  | 2017-03-06   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.3.1   | `jessie`  | 2016-10-31   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.3.0   | `jessie`  | 2016-06-21   | monthly          | `r-ver`
3.2.0   | `jessie`  | 2015-06-18   | monthly          | `r-ver`
3.1.0   | `jessie`  | 2014-09-17   | monthly          | `r-ver`

@@

### Maintenance and Updates

@@

### License

@@

# Conclusions

Over the past several years, Docker has seen immense adoption across industry and academia. 
The Open Container initiative \citep{oci} now provides an open standard that has further
extended this container approach to research environments through projects such as 
Singularity \citep{singularity}, allowing users to deploy containerized environments such as Rocker on machines where they do not have root access, such as clusters or private servers. Containerization promises
to solve numerous challenges such as portability and replicability in research computing,
which often relies on complex and heterogeneous software stacks \citep{Boettiger2015}. Yet
implementing such environments in containers is not a trivial task, and not all implementations
provide the same usability, portability or reproducibility.  Here we have detailed the 
approach taken by the Rocker project in creating and maintaining these environments through
an open and community-driven process.  This structure of the Rocker project has evolved
over three years of operation while drawing in an ever widening base of academic researchers, 
university instructors and industry users.  We believe this overview will be instructive
not only to users and developers interested in the Rocker project, but as a model for
similar efforts around other environments or domains.
